Gene Enrichment Analysis with MSigDB

### 2022-07-01 MSigDB.Rmd
fixing problem with order of tes v. target

all_genes = msigdbr(species = "Rattus norvegicus")

@@

setwd("C:/Users/angel/OneDrive/Desktop/Saba Lab/Data/")
#setwd("C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data")

@@

@@

# getting entrez ids
get_entrez = getBM(attributes=c("ensembl_gene_id",
                              "external_gene_name",
                              "description",
                              "chromosome_name",
                              "start_position",
                              "end_position",
                              "gene_biotype",
                            #"entrezgene_trans_name",
                            "entrezgene_id"),
                 filters = 'biotype',
                 values = "protein_coding",
                 mart = ensembl)

@@

missing_b = combine.brain[combine.brain %!in% entrez_brain$ensembl_gene_id]
# 15 missing TES : "ENSRNOG00000047739" "ENSRNOG00000008563" "ENSRNOG00000009140" "ENSRNOG00000038437" "ENSRNOG00000034130" "ENSRNOG00000003887" "ENSRNOG00000007374" "ENSRNOG00000013441" "ENSRNOG00000003891" "ENSRNOG00000038980" "ENSRNOG00000013228" "ENSRNOG00000038150" "ENSRNOG00000049171" "ENSRNOG00000011074" "ENSRNOG00000050172"

@@

missing_l = combine.liver[combine.liver %!in% entrez_liver$ensembl_gene_id]
# 12 missing TES : "ENSRNOG00000020805" "ENSRNOG00000050499" "ENSRNOG00000049727" "ENSRNOG00000033130" "ENSRNOG00000042543" "ENSRNOG00000046529" "ENSRNOG00000013464" "ENSRNOG00000049515" "ENSRNOG00000013410" "ENSRNOG00000016535" "ENSRNOG00000033619" "ENSRNOG00000033245"

@@

@@

@@

setwd("C:/Users/angel/OneDrive/Desktop/Saba Lab/Data/")
#setwd("C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data")
entrez_brain = read.csv("EntrezIds_brain.csv")
entrez_liver = read.csv("EntrezIds_liver.csv")

@@

msigdbr_t2g = all_genes %>% dplyr::distinct(gs_name, entrez_gene) %>% as.data.frame()

@@

# brain.list, liver.list

out.brain = list(c(), c(), c(), c(), c())

for (i in 1:5)
{
  vect = brain.list[[i]]
  entrez = entrez_brain$entrezgene_id[entrez_brain$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = msigdbr_t2g, pAdjustMethod = "fdr")
  out.brain[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

out.liver = list(c(), c(), c(), c(), c())

for (i in 1:5)
{
  vect = liver.list[[i]]
  entrez = entrez_liver$entrezgene_id[entrez_liver$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = msigdbr_t2g, pAdjustMethod = "fdr")
  out.liver[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

brain.ids = c(paste(out.brain[[1]]$ID, collapse = ", "), paste(out.brain[[2]]$ID, collapse = ", "), paste(out.brain[[3]]$ID, collapse = ", "), paste(out.brain[[4]]$ID, collapse = ", "), paste(out.brain[[5]]$ID, collapse = ", "))
names(brain.ids) = top.brain[,1]

knitr::kable(brain.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Brain", colnames = "")

liver.ids = c(paste(out.liver[[1]]$ID, collapse = ", "), paste(out.liver[[2]]$ID, collapse = ", "), paste(out.liver[[3]]$ID, collapse = ", "), paste(out.liver[[4]]$ID, collapse = ", "), paste(out.liver[[5]]$ID, collapse = ", "))
names(liver.ids) = top.liver[,1]

knitr::kable(liver.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Liver", colnames = "")

@@

>>>>

H_genes = msigdbr(species = "Rattus norvegicus", category = "H")
C2_genes = msigdbr(species = "Rattus norvegicus", category = "C2")
C5_genes = msigdbr(species = "Rattus norvegicus", category = "C5")
C8_genes = msigdbr(species = "Rattus norvegicus", category = "C8")

@@

#setwd("C:/Users/angel/OneDrive/Desktop/Saba Lab/Data/")
setwd("C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data")

@@

When retrieving Entrez IDs for our gene sets of interest, I found that many of the genes within these sets were no longer available within biomaRt. Looking at the Ensembl.org website, many of these ensembl IDs were retired via Ensembl. The version in which these ensembl IDs were available is version 104. This version was used for the enrichment analysis. However, there may be concerns using these retired ensembl IDs, especially where there is no longer a corresponding ID in the new version. Below is a table summarizing the number of genes missing, along with examples of pages from defunct ensembl ID webpages.

@@

# getting entrez ids
get_entrez = getBM(attributes=c("ensembl_gene_id",
                              "external_gene_name",
                              "description",
                              "chromosome_name",
                              "start_position",
                              "end_position",
                              "gene_biotype",
                            #"entrezgene_trans_name",
                            "entrezgene_id"),
                 filters = 'biotype',
                 values = "protein_coding",
                 mart = ensembl.miss)

@@

missing_b = combine.brain[combine.brain %!in% entrez_brain.1$ensembl_gene_id]
# 15 missing TES : "ENSRNOG00000047739" "ENSRNOG00000008563" "ENSRNOG00000009140" "ENSRNOG00000038437" "ENSRNOG00000034130" "ENSRNOG00000003887" "ENSRNOG00000007374" "ENSRNOG00000013441" "ENSRNOG00000003891" "ENSRNOG00000038980" "ENSRNOG00000013228" "ENSRNOG00000038150" "ENSRNOG00000049171" "ENSRNOG00000011074" "ENSRNOG00000050172"

@@

missing_l = combine.liver[combine.liver %!in% entrez_liver.1$ensembl_gene_id]
# 12 missing TES : "ENSRNOG00000020805" "ENSRNOG00000050499" "ENSRNOG00000049727" "ENSRNOG00000033130" "ENSRNOG00000042543" "ENSRNOG00000046529" "ENSRNOG00000013464" "ENSRNOG00000049515" "ENSRNOG00000013410" "ENSRNOG00000016535" "ENSRNOG00000033619" "ENSRNOG00000033245"

@@

missingness = data.frame(dataset = c("TES Brain", "Sig Target Genes Brain", "TES Liver", "Sig Target Genes Liver"), missing = c(15, 28, 12, 116), total = c(174, 138, 35, 930))

knitr::kable(missingness, caption = "Missing Ensembl IDs within Datasets")

![Missing Gene: ENSRNOG00000050885](C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data/ENSRNOG00000050885_Ensembl.png)

![Missing Gene: ENSRNOG00000013464](C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data/ENSRNOG00000013464_Ensembl.png)

@@

## Enrichment Analysis

@@

#setwd("C:/Users/angel/OneDrive/Desktop/Saba Lab/Data/")
setwd("C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data")
entrez_brain = read.csv("EntrezIds_brain.csv")
entrez_liver = read.csv("EntrezIds_liver.csv")

@@

msigdbr_t2g_H = H_genes %>% dplyr::distinct(gs_name, entrez_gene) %>% as.data.frame()
msigdbr_t2g_C2 = C2_genes %>% dplyr::distinct(gs_name, entrez_gene) %>% as.data.frame()
msigdbr_t2g_C5 = C5_genes %>% dplyr::distinct(gs_name, entrez_gene) %>% as.data.frame()
msigdbr_t2g_C8 = C8_genes %>% dplyr::distinct(gs_name, entrez_gene) %>% as.data.frame()

@@

# H

out.brain_H = list(c(), c(), c(), c(), c())
database = msigdbr_t2g_H

for (i in 1:5)
{
  vect = brain.list[[i]]
  entrez = entrez_brain$entrezgene_id[entrez_brain$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = database, pAdjustMethod = "fdr")
  out.brain_H[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

out.liver_H = list(c(), c(), c(), c(), c())

for (i in 1:5)
{
  vect = liver.list[[i]]
  entrez = entrez_liver$entrezgene_id[entrez_liver$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = database, pAdjustMethod = "fdr")
  out.liver_H[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

#brain.ids = c(paste(out.brain[[1]]$ID, collapse = ", "), paste(out.brain[[2]]$ID, collapse = ", "), paste(out.brain[[3]]$ID, collapse = ", "), paste(out.brain[[4]]$ID, collapse = ", "), paste(out.brain[[5]]$ID, collapse = ", "))
#names(brain.ids) = top.brain[,1]

#knitr::kable(brain.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Brain", colnames = "")

#liver.ids = c(paste(out.liver[[1]]$ID, collapse = ", "), paste(out.liver[[2]]$ID, collapse = ", "), paste(out.liver[[3]]$ID, collapse = ", "), paste(out.liver[[4]]$ID, collapse = ", "), paste(out.liver[[5]]$ID, collapse = ", "))
#names(liver.ids) = top.liver[,1]

#knitr::kable(liver.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Liver", colnames = "")

# C2

out.brain_C2 = list(c(), c(), c(), c(), c())
database = msigdbr_t2g_C2

for (i in 1:5)
{
  vect = brain.list[[i]]
  entrez = entrez_brain$entrezgene_id[entrez_brain$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = database, pAdjustMethod = "fdr")
  out.brain_C2[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

out.liver_C2 = list(c(), c(), c(), c(), c())

for (i in 1:5)
{
  vect = liver.list[[i]]
  entrez = entrez_liver$entrezgene_id[entrez_liver$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = database, pAdjustMethod = "fdr")
  out.liver_C2[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

#brain.ids = c(paste(out.brain[[1]]$ID, collapse = ", "), paste(out.brain[[2]]$ID, collapse = ", "), paste(out.brain[[3]]$ID, collapse = ", "), paste(out.brain[[4]]$ID, collapse = ", "), paste(out.brain[[5]]$ID, collapse = ", "))
#names(brain.ids) = top.brain[,1]

#knitr::kable(brain.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Brain", colnames = "")

#liver.ids = c(paste(out.liver[[1]]$ID, collapse = ", "), paste(out.liver[[2]]$ID, collapse = ", "), paste(out.liver[[3]]$ID, collapse = ", "), paste(out.liver[[4]]$ID, collapse = ", "), paste(out.liver[[5]]$ID, collapse = ", "))
#names(liver.ids) = top.liver[,1]

#knitr::kable(liver.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Liver", colnames = "")

# C5

out.brain_C5 = list(c(), c(), c(), c(), c())
database = msigdbr_t2g_C5

for (i in 1:5)
{
  vect = brain.list[[i]]
  entrez = entrez_brain$entrezgene_id[entrez_brain$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = database, pAdjustMethod = "fdr")
  out.brain_C5[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

out.liver_C5 = list(c(), c(), c(), c(), c())

for (i in 1:5)
{
  vect = liver.list[[i]]
  entrez = entrez_liver$entrezgene_id[entrez_liver$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = database, pAdjustMethod = "fdr")
  out.liver_C5[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

#brain.ids = c(paste(out.brain[[1]]$ID, collapse = ", "), paste(out.brain[[2]]$ID, collapse = ", "), paste(out.brain[[3]]$ID, collapse = ", "), paste(out.brain[[4]]$ID, collapse = ", "), paste(out.brain[[5]]$ID, collapse = ", "))
#names(brain.ids) = top.brain[,1]

#knitr::kable(brain.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Brain", colnames = "")

#liver.ids = c(paste(out.liver[[1]]$ID, collapse = ", "), paste(out.liver[[2]]$ID, collapse = ", "), paste(out.liver[[3]]$ID, collapse = ", "), paste(out.liver[[4]]$ID, collapse = ", "), paste(out.liver[[5]]$ID, collapse = ", "))
#names(liver.ids) = top.liver[,1]

#knitr::kable(liver.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Liver", colnames = "")

# C8

out.brain_C8 = list(c(), c(), c(), c(), c())
database = msigdbr_t2g_C8

for (i in 1:5)
{
  vect = brain.list[[i]]
  entrez = entrez_brain$entrezgene_id[entrez_brain$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = database, pAdjustMethod = "fdr")
  out.brain_C8[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

out.liver_C8 = list(c(), c(), c(), c(), c())

for (i in 1:5)
{
  vect = liver.list[[i]]
  entrez = entrez_liver$entrezgene_id[entrez_liver$ensembl_gene_id %in% vect]
  entrez = na.omit(entrez)
  listt = enricher(gene = entrez, TERM2GENE = database, pAdjustMethod = "fdr")
  out.liver_C8[[i]] = data.frame(ID = listt$ID, Description = listt$Description, pvalue = listt$pvalue, p.adjust = listt$p.adjust, geneID = listt$geneID)
}

#brain.ids = c(paste(out.brain[[1]]$ID, collapse = ", "), paste(out.brain[[2]]$ID, collapse = ", "), paste(out.brain[[3]]$ID, collapse = ", "), paste(out.brain[[4]]$ID, collapse = ", "), paste(out.brain[[5]]$ID, collapse = ", "))
#names(brain.ids) = top.brain[,1]

#knitr::kable(brain.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Brain", colnames = "")

#liver.ids = c(paste(out.liver[[1]]$ID, collapse = ", "), paste(out.liver[[2]]$ID, collapse = ", "), paste(out.liver[[3]]$ID, collapse = ", "), paste(out.liver[[4]]$ID, collapse = ", "), paste(out.liver[[5]]$ID, collapse = ", "))
#names(liver.ids) = top.liver[,1]

#knitr::kable(liver.ids, caption = "Enrichment of Target Liver Genes Influenced by TES Protein-Coding Genes from the Liver", colnames = "")

first.brain = rbind(out.brain_H[[1]], out.brain_C2[[1]], out.brain_C5[[1]], out.brain_C8[[1]])
second.brain = rbind(out.brain_H[[2]], out.brain_C2[[2]], out.brain_C5[[2]], out.brain_C8[[2]])
third.brain = rbind(out.brain_H[[3]], out.brain_C2[[3]], out.brain_C5[[3]], out.brain_C8[[3]])
fourth.brain = rbind(out.brain_H[[4]], out.brain_C2[[4]], out.brain_C5[[4]], out.brain_C8[[4]])
fifth.brain = rbind(out.brain_H[[5]], out.brain_C2[[5]], out.brain_C5[[5]], out.brain_C8[[5]])

first.liver = rbind(out.liver_H[[1]], out.liver_C2[[1]], out.liver_C5[[1]], out.liver_C8[[1]])
second.liver = rbind(out.liver_H[[2]], out.liver_C2[[2]], out.liver_C5[[2]], out.liver_C8[[2]])
third.liver = rbind(out.liver_H[[3]], out.liver_C2[[3]], out.liver_C5[[3]], out.liver_C8[[3]])
fourth.liver = rbind(out.liver_H[[4]], out.liver_C2[[4]], out.liver_C5[[4]], out.liver_C8[[4]])
fifth.liver = rbind(out.liver_H[[5]], out.liver_C2[[5]], out.liver_C5[[5]], out.liver_C8[[5]])

#setwd("C:/Users/angel/OneDrive/Desktop/Saba Lab/Data/")
setwd("C:/Users/yodeange/OneDrive - The University of Colorado Denver/Saba Lab/Data")
#write.csv(first.brain, "Brain Enrichment_1.csv")
#write.csv(second.brain, "Brain Enrichment_2.csv")
#write.csv(third.brain, "Brain Enrichment_3.csv")
#write.csv(fourth.brain, "Brain Enrichment_4.csv")
#write.csv(fifth.brain, "Brain Enrichment_5.csv")

#write.csv(first.liver, "liver Enrichment_1.csv")
#write.csv(second.liver, "liver Enrichment_2.csv")
#write.csv(third.liver, "liver Enrichment_3.csv")
#write.csv(fourth.liver, "liver Enrichment_4.csv")
#write.csv(fifth.liver, "liver Enrichment_5.csv")

@@

### Enrichment of Target Liver Protein-Coding Genes acted upon by TES Protein-Coding Genes from Brain

knitr::kable(first.brain, caption = "Brain TES Protein-Coding Gene: ENSRNOG00000000010")
knitr::kable(second.brain, caption = "Brain TES Protein-Coding Gene: ENSRNOG00000020519")
knitr::kable(third.brain, caption = "Brain TES Protein-Coding Gene: ENSRNOG00000002381")
knitr::kable(fourth.brain, caption = "Brain TES Protein-Coding Gene: ENSRNOG00000016567")
knitr::kable(fifth.brain, caption = "Brain TES Protein-Coding Gene: ENSRNOG00000008176")

### Enrichment of Target Brain Protein-Coding Genes acted upon by TES Protein-Coding Genes from Liver

knitr::kable(first.liver, caption = "Liver TES Protein-Coding Gene: ENSRNOG00000032737")
knitr::kable(second.liver, caption = "Liver TES Protein-Coding Gene: ENSRNOG00000021029")
knitr::kable(third.liver, caption = "Liver TES Protein-Coding Gene: ENSRNOG00000008422")
knitr::kable(fourth.liver, caption = "Liver TES Protein-Coding Gene: ENSRNOG00000002964")
knitr::kable(fifth.liver, caption = "Liver TES Protein-Coding Gene: ENSRNOG00000011973")
